### 2018-03-15 paper/3-abstractions.tex

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: Store i k v -> i -> Store i k v
getValue   :: Store i k v -> k -> v
putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
getHash    :: Hashable v => Store i k v -> k -> Hash
initialise :: Hashable v => i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- State monad
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{-2mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We begin by establishing a common vocabulary for build systems.
% that allows
% us to abstract away from specific application domains, such as software
% development or spreadsheets.
Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
and gives types of the library functions we use in this paper.

@@

Now we can see why the \hs{Task} abstraction needed to be abstracted over \hs{f}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
The \hs{busy} build system instantiates \hs{c} to \hs{Monad} and \hs{f} to \hs{State (Store i k v)},
so that the call-back to \hs{fetch :: k -> f v} can side-effect the \hs{Store},
thereby allowing successive calls to \hs{fetch} to communicate with one another.
We will see why we want \hs{Task} to be \emph{polymorphic} in \hs{f} in \S\ref{sec-deps}.

@@

An alert reader might notice that \hs{busy} accepts \emph{monadic tasks} of
type \hs{Task}~\hs{Monad}, whereas \hs{sprsh1} is a \hs{Task}~\hs{Applicative}.
This indicates that \hs{busy} can cope with tasks containing dynamic
dependencies, which we discuss next (\S\ref{sec-task-monad}).

It is interesting to note that if \hs{busy} is given a \emph{functorial task}
of type \hs{Task}~\hs{Functor}, which cannot use the application operator
\hs{<*>}, then \hs{busy} can rejoice in knowing that it will never have to
rebuild a key twice. Indeed, functorial tasks are very much like tail recursive
functions -- they cannot branch into multiple subproblems, always leading to a
straight-line build.

@@

As explained in~\S\ref{sec-background-excel}, some task descriptions have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such task descriptions correspond to the type
\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
example:

@@

\noindent
The main difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
then \hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.

@@

To express this formally we need an auxiliary
function \hs{recompute}, that computes the value of a key in a given store
\emph{without attempting to update any dependencies}:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
newtype Identity a = Identity { runIdentity :: a }
\end{minted}
\vspace{0mm}
\begin{minted}[xleftmargin=10pt]{haskell}
recompute :: Task Monad k v -> Store i k v -> k -> Maybe v
recompute task store key = runIdentity <$> task (Identity . getValue store) key
\end{minted}
\vspace{1mm}
\noindent
\todo{SLPJ}{Why monad?}
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad.
The use of \hs{Identity} just fixes the `impedance mismatch'
between the pure function \hs{getValue}~\hs{store}, whose type is \store, and
the \hs{fetch} argument of the \hs{task}, whose type must be
\hs{k}~\hs{->}~\hs{f}~\hs{v} for some \hs{f}. To fix the mismatch, we wrap the
result of the pure function in the \hs{Identity} monad: the function
\hs{Identity}~\hs{.}~\hs{getValue}~\hs{store} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now need to get rid
of the \hs{Identity} wrapper by applying \hs{runIdentity} to the contents of
\hs{Maybe}.

@@

...

We can statically analyse an applicative task and extract all dependencies of a
key using the following function:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Data.Functor.Const
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task = maybe [@@] getConst . task (\k -> Const [k])
\end{minted}
\vspace{1mm}

\noindent
We can use this function to extract dependencies of \hs{sprsh1}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

\noindent
It is important to note that we extract the dependencies without supplying any
values to the build task. This guarantees that no actual computation is performed
during the analysis -- we cannot execute the function \hs{(+)} if we have no
values for it! Static analysis of applicative computations is not new and has
been used to similar effect before, e.g. see~\cite{free-applicatives}.

To understand how \hs{dependencies} works, let us examine its type more closely,
expanding the type synonym \hs{Task}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
(@\std{forall}@ f. Applicative f => (k -> f v) -> k -> Maybe (f v)) -> k -> [k]
\end{minted}
\vspace{1mm}

\noindent
Since the first argument is polymorphic in \hs{f}, we can choose any \hs{f} that
satisfies the \hs{Applicative} constraint. A good choice in this case is the
applicative functor \hs{Const}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
newtype Const m a = Const { getConst :: m }
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{1mm}

\noindent
A value of type \hs{Const}~\hs{[}\hs{k]}~\hs{v} contains no values \hs{v} but
allows us to record all invocations of the custom \hs{fetch} function that we
feed to the given compute: \hs{fetch k = Const [k]}. The compute returns a
value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)} that can be
unwrapped using \hs{maybe [@@] getConst}, i.e. returning the empty list of
dependencies in case of an input.

To extract the dependencies of monadic tasks we introduce the function \hs{track}:
a combination of \hs{recompute} and \hs{dependencies} that takes a store as input
and computes both the resulting value and the list of its dependencies:

@@

\noindent
As expected, the dependencies of \hs{sprsh2} are determined by the contents of
the store.

...

>>>>

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: Store i k v -> i -> Store i k v
getValue   :: Store i k v -> k -> v
putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
getHash    :: Hashable v => Store i k v -> k -> Hash
initialise :: Hashable v => i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad (from Control.Monad.State)
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
\end{minted}
\begin{minted}[fontsize=\small]{haskell}
newtype Const ks a = Const { getConst :: ks }
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const ks) where
    fmap _ (Const ks) = Const ks
instance Monoid ks => Applicative (Const ks) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{-2mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We begin by establishing a common vocabulary for build systems.
% that allows
% us to abstract away from specific application domains, such as software
% development or spreadsheets.
Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
and gives types of the library functions we use in this paper.

@@

@@

\subsection{The need for polymorphism in \hs{Task}}

Now we can see why the \hs{Task} abstraction is polymorphic in \hs{f}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
The \hs{busy} build system instantiates \hs{f} to \hs{State (Store i k v)},
so that the call-back to \hs{fetch :: k -> f v} can side-effect the \hs{Store},
thereby allowing successive calls to \hs{fetch} to communicate with one another.

We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
Given \emph{one} task description T, we want to explore \emph{many} build systems that
can build T -- and we will do so in Sections~\ref{sec-build} and \ref{sec-implementations}.
As we shall see, each build system will use a different \hs{f}, so the task description
must not fix \hs{f}.

But nor can the task description work for \emph{any} \hs{f}; most
task descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that
\hs{f} satisfies certain properties, such as \hs{Monad} or
\hs{Applicatove}.  That is why \hs{task} has the ``\hs{c f =>}''
constraint in its type, expressing that \hs{f} can only be instantiated by types
that satisfy the constraints \hs{c}.

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emgerged, we find the that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task Applicative}. In \hs{sprsh1} we needed only \hs{Applicative} operations,
  expressing the fact that the dependencies between cells can be determined \emph{statically};
  that is, by looking at the formulae, without ``executing'' them (see \S\ref{sec-deps}).
\item \hs{Task Monad}.  As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula
  may depend on the value of cell C, but \emph{which} cell C depends on
  the value of another cell D.
\item \hs{Task Functor} is somehat degenerate: the task description cannot even use
  the application operator \hs{<*>}, which limits dependencies to a single linear chain.
\end{itemize}

Notice also that, even though \hs{busy} takes a \hs{Task Monad} as its argument, an application of
\hs{busy} to a \hs{Task Applicative} will typecheck just fine.  It feels a bit
like sub-typing, but it is actually just ordinary higher-rank polymorphism at work.

@@

As explained in~\S\ref{sec-background-excel}, some task descriptions have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such task descriptions correspond to the type
\hs{Task Monad k v}. Consider this spreadsheet
example:

@@

\noindent
The big difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
then \hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.

@@

To express this formally we need an auxiliary
function \hs{recompute}, that computes the value of a key in a given store
\emph{without attempting to update any dependencies}:
\begin{minted}[xleftmargin=10pt]{haskell}
recompute :: Task Monad k v -> Store i k v -> k -> Maybe v
recompute task store key = runIdentity <$> task (Identity . getValue store) key
\end{minted}
\noindent
\todo{SLPJ}{Why monad?}
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad.
The use of \hs{Identity} just fixes the `impedance mismatch'
between the pure function \hs{getValue}~\hs{store}, whose type is \store, and
the \hs{fetch} argument of the \hs{task}, whose type must be
\hs{k}~\hs{->}~\hs{f}~\hs{v} for some \hs{f}. To fix the mismatch, we wrap the
result of the pure function in the \hs{Identity} monad: the function
\hs{Identity}~\hs{.}~\hs{getValue}~\hs{store} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now need to get rid
of the \hs{Identity} wrapper by applying \hs{runIdentity} to the contents of
\hs{Maybe}.

@@

Earlier we remarked that a \hs{Task Applicative} could only have static dependencies.
Usually we would extract such static dependencies by (in the case of Excel)
looking at the syntax tree of the formula.  But a task description has no
such syntax tree.  Yet, remarkably, we can use the polymorphism of a
\hs{Task Applicative} to find its dependencies \emph{without dong any of the actual work}.
Here is the code:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task k = case task (\k -> Const [k]) k of
                        Nothing         -> []
                        Just (Const ks) -> ks
\end{minted}
\vspace{1mm}
Here \hs{Const} is a standard Haskell type defined in Figure~\ref{fig-types}.
We instantiate \hs{f} to \hs{Const [k]}.  So a value of type \hs{f v},
or in this case \hs{Const [k] v}, contains no value \hs{v}, but does contain
a list of keys of type \hs{[k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task}
records a single dependency; and the standard definitions of \hs{Applicative}
for \hs{Const} (which we give in Figure~\ref{fig-types}) combines the
dependencies from different parts of the task.
As a result, running the task with \hs{f} = \hs{Const [k]} will accumulate a list of
the task's dependencies -- and that is just what \hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

\noindent
Notice that these calls to \hs{dependencies} do no actual computation
(in this case, spreadsheet arithmetic). They can't:
we are not supplying a store or any input numbers.
So, through the wonders of polymorphism, we are able to extract the
dependencies of the spreadsheet formula, and to do so efficiently,
simply by running its code in a different \hs{Applicative}!
This is not new -- e.g. see~\cite{free-applicatives} -- but it is cool.

So much for applicatives.  What about monadic tasks with dynamic dependencies?
We introduce the function \hs{track}:
a combination of \hs{recompute} and \hs{dependencies} that takes a store as input
and computes both the resulting value and the list of its dependencies:

@@

\noindent
As expected, the dependencies of \hs{sprsh2} are determined by the contents of
the store.
\todo{slpj}{This track stuff is not well motivated. When do we want it?  How do we use it?}

\todo{slpj}{I think we should push 'inputs' into the appendix.}
